Concurrent Programming with CXL

### 2020-06-24 book/paper.tex
Finished a draft of the non-blocking synchronization chapter

@@

While model checking does not in general prove an algorithm correct,
it can help with proving an algorithm correct.
The reason is that the correctness of many programs is built around
\emph{invariants}:
\index{invariant}
\glsadd{invariant}
predicates that must hold for every state in the
execution of a program.  A model checker can find violations of
proposed invariants when evaluating a model and provide valuable early
feedback to somebody who is trying to construct a proof, even an
informal one.
We will include examples
of such invariants as they often provide excellent insight into
why a particular algorithm works.

@@

Introduced by Dijkstra,
a \emph{semaphore} is a synchronization primitive that fits the bill~\cite{EWD35}.
A semaphore is essentially
a counter that can be incremented and decremented but is not allowed to go
below zero.  The semaphore counter is typically initialized to the number of
units of a resource initially available.
When allocating a resource, a process decrements the
counter using the \texttt{P}
\index{P}
\index{Procure}
operation.  You can think of \texttt{P} standing
for \emph{Procure}, as in procuring the resource associated with the semaphore.
The \texttt{P} operation blocks the invoking process if the counter is zero.
To release the resource, a process increments the resource using the
\texttt{V}
\index{V}
\index{Vacate}
operation.  You can think of \texttt{V} standing for \emph{vacate},
as in vacating the resource.
One can think of a semaphore as a generalization of a lock.  In fact, a
lock can be implemented by a semaphore initialized to 1.  Then \texttt{P}
procures the lock, and \texttt{V} vacates the lock.

@@

\section*{Exercises}
\begin{problems}
\item Figure~\ref{fig:bank} shows an implementation of a bank with various
accounts and transfers between those accounts.
Unfortunately, running the test reveals that it sometimes leaves unterminated
processes.  Can you fix the problem?
\item Add a method \texttt{total()} to the solution of the previous question
that computes the total over all balances.
It needs to obtain a lock on all accounts.  Make sure that
it cannot cause deadlock.
\item Implement an extra process that checks, once, if the total of
the balances is the same as that at the beginning (using an assertion).
Is once enough to determine that no money gets lost or create, or should it do
so continually?
\end{problems}

@@

So far we have concentrated on critical sections to synchronize multiple
processes.  Certainly, preventing multiple processes from accessing
certain code at the same time simplifies how to think about synchronization.
However, it can lead to starvation.  Even in the absence of starvation,
if some process is slow for some reason while being in the critical section,
the other processes have to wait for it to finish the critical section.
In this chapter, we will have a brief look at how one can develop concurrent
code in which processes never have to wait for other processes.
For this, we will revisit the producer/consumer problem, in which a
set of processes are producing items and a set of processes are
consuming items.

The code in Figure~\ref{fig:hw} is based on code developed by Herlihy and
Wing~\cite{HW87}.
The code is a ``proof of existence'' for non-blocking synchronization; it
is not necessarily practical.
There are two variables.  \texttt{items} is an unbounded array with each
entry initialized to \texttt{()}.  \texttt{back} is an index into the
array and points to the next slot where a new value is inserted.
The code uses two interlock instructions:
\begin{itemize}
\item \texttt{inc(p)}: atomically increments \texttt{\^{}p} and returns
the old value;
\item \texttt{swap(p)}: sets \texttt{\^{}p} to \texttt{()} and returns
the old value.
\end{itemize}

@@

There is no critical section.  If one process is executing instructions
very slowly, this does not negatively impact the other processes, as it
would with solutions based on critical sections.
On the contrary, it helps them because it creates less contention.

The solution is not practical because the \texttt{items} array must be
unbounded if an unbounded number of items may be produced.  Each slot
in the array is only used once, which is inefficient.  Also, the \texttt{inc}
and \texttt{swap} interlock instructions are not universally available
on existing hardware.
However, in the literature there are many examples of practical
non-blocking (aka \emph{wait-free})
\index{wait-free synchronization}
synchronization algorithms~\cite{}.
There are also various algorithms where read-only access does not
require a lock but updates do require a lock---these are useful in
situations where most accesses are read-only.

TODO: Talk about seqlock perhaps.

@@

>>>>

\newglossaryentry{linearizable}
  name=linearizable,
  description={is a consistency condition for concurrent access to an object, requiring
    that each access must appear to execute atomically sometime between the invocation
    of the access and its completion}

@@

\glsadd{invariant}
While model checking does not in general prove an algorithm correct,
it can help with proving an algorithm correct.
The reason is that the correctness of many programs is built around
\emph{invariants}:
\index{invariant}
predicates that must hold for every state in the
execution of a program.  A model checker can find violations of
proposed invariants when evaluating a model and provide valuable early
feedback to somebody who is trying to construct a proof, even an
informal one.
We will include examples
of such invariants as they often provide excellent insight into
why a particular algorithm works.

@@

Introduced by Dijkstra,
a \emph{semaphore} is a synchronization primitive that fits the bill~\cite{EWD35}.
A semaphore is essentially
a counter that can be incremented and decremented but is not allowed to go
below zero.  The semaphore counter is typically initialized to the number of
units of a resource initially available.
When allocating a resource, a process decrements the
counter using the \texttt{P}
\index{P}
\index{procure}
operation.  You can think of \texttt{P} standing
for \emph{Procure}, as in procuring the resource associated with the semaphore.
The \texttt{P} operation blocks the invoking process if the counter is zero.
To release the resource, a process increments the resource using the
\texttt{V}
\index{V}
\index{vacate}
operation.  You can think of \texttt{V} standing for \emph{vacate},
as in vacating the resource.
One can think of a semaphore as a generalization of a lock.  In fact, a
lock can be implemented by a semaphore initialized to 1.  Then \texttt{P}
procures the lock, and \texttt{V} vacates the lock.

@@

\section*{Exercises}
\begin{problems}
\item \label{ex:bank} Figure~\ref{fig:bank} shows an implementation of a bank with various
accounts and transfers between those accounts.
Unfortunately, running the test reveals that it sometimes leaves unterminated
processes.  Can you fix the problem?
\item Add a method \texttt{total()} to the solution of the previous question
that computes the total over all balances.
It needs to obtain a lock on all accounts.  Make sure that
it cannot cause deadlock.
\item Implement an extra process that checks, once, if the total of
the balances is the same as that at the beginning (using an assertion).
Is once enough to determine that no money gets lost or create, or should it do
so continually?
\end{problems}

@@

So far we have concentrated on critical sections to synchronize multiple
processes.  Certainly, preventing multiple processes from accessing
certain code at the same time simplifies how to think about synchronization.
However, it can lead to starvation.  Even in the absence of starvation,
if some process is slow for some reason while being in the critical section,
the other processes have to wait for it to finish executing the critical section.
In this chapter, we will have a look at how one can develop concurrent
code in which processes do not have to wait for other processes.

As a first example, we will revisit the producer/consumer problem.
The code in Figure~\ref{fig:hw} is based on code developed by Herlihy and
Wing~\cite{HW87}.
The code is a ``proof of existence'' for non-blocking synchronization; it
is not necessarily practical.
There are two variables.  \texttt{items} is an unbounded array with each
entry initialized to \texttt{()}.  \texttt{back} is an index into the
array and points to the next slot where a new value is inserted.
The code uses two interlock instructions:
\begin{itemize}
\item \texttt{inc(p)}: atomically increments \texttt{\^{}p} and returns
the old value;
\item \texttt{swap(p)}: sets \texttt{\^{}p} to \texttt{()} and returns
the old value.
\end{itemize}

@@

There is no critical section.  If one process is executing instructions
very slowly, this does not negatively impact the other processes, as it
would with solutions based on critical sections.
On the contrary, it helps them because it creates less contention.
Unfortunately, the solution is not practical for the following reasons:
\begin{itemize}
\item The \texttt{items} array must be of infinite size if an unbounded number
of items may be produced;
\item Each slot in the array is only used once, which is inefficient;
\item the \texttt{inc} and \texttt{swap} interlock instructions are not
universally available on existing processors.
\end{itemize}
However, in the literature there are many examples of practical
non-blocking (aka \emph{wait-free})
\index{wait-free synchronization}
synchronization algorithms~\cite{}.

There are also various algorithms where read-only access is wait-free
but updates do require a lock---these are useful in situations where
most accesses are read-only and the performance of updates is less
critical~\cite{}.
Figures~\ref{fig:lst1} and~\ref{fig:lst2} implements an ordered linked
list of integers without duplicates.
There are two update operations:
values can be added using \texttt{lst\_insert} or deleted using
\texttt{lst\_remove}.
There is also a read-only operation: \texttt{lst\_contains} checks if
a particular value is in the list.

\glsadd{linearizable}
The implementation uses the \texttt{alloc} module for dynamic allocation
\index{dynamic allocation}
of nodes in the list using \texttt{recAlloc} and \texttt{recFree}.
The read-only operation \texttt{lst\_contains} is wait\_free: it scans
through the list without having to obtain a lock.
Nonetheless, the implementation of the list is
\emph{linearizable}~\cite{HW90},
\index{linearizable}
a strong notion of consistency that
makes it appear as if each of the operations executes atomically at
some point between their invocation and return.

The list has two ``book-end'' nodes with values \texttt{-inf} and
\texttt{inf} (like the Python \texttt{math.inf} constants).
Each node has a lock, a value, and \texttt{next}, a pointer to the next node
(which is \texttt{NULL} for the final \texttt{inf} node).
The \texttt{lst\_find(lst, v)} method is a helper function for update
methods that
finds and locks two consecutive nodes \texttt{before} and \texttt{after}
such that \texttt{before.value $<$ v $\le$ after.value}.
It does so by performing something called \emph{hand-over-hand locking}.
\index{hand-over-hand locking}
It first locks the first node, which is the \texttt{-inf} node.
Then, iteratively, it obtains a lock on the next node and release the
lock on the last one, and so on, similar to climbing a tree
hand-over-hand.

Having the two adjoining nodes locked, implementing \texttt{lst\_insert}
and \texttt{lst\_remove} is straightforward.
The \texttt{lst\_contains} method can simply scan through the list
because the \texttt{next} pointers are updated atomically in such a
way that they always point to a legal suffix of the list.

@@

\section*{Exercises}
\begin{problems}
\item A \emph{seqlock} consists of a lock and a version number.
An update operation acquires the lock, increments the version number, makes the
changes to the data structure, and then releases the lock.  A read-only operation
does not use the lock.  Instead, it retrieves the version number,
reads the data structure, and then checks if the
version number has not changed.  If it has, read-only operation is retried.
Use a seqlock to implement a bank much like Exercise~\ref{ex:bank}, with
one seqlock for the entire bank (i.e., no locks on individual accounts).
Method \texttt{transfer} is an update operation; method \texttt{total} is a
read-only operation.  Explain how a \texttt{seqlock} can lead to starvation.
\end{problems}
